---
title: "More Linear Regression"
subtitle: "Stat 243: Lab 3"
author: "Insert Your Name Here"
format: pdf
editor: source
---

## Due: Friday, Oct 2nd, 11am PT.

## Instructions

Problem sets are a (nearly) weekly opportunity to engage with ideas relevant to content taught so far in the course. As your are working through your problem set, make sure to "Render" the document early and often. Take a look at the PDF output, as this is what you will turn in on Gradescope. Make sure the output is fully readable and is something that you are proud of. Take care to make sure code, writing, and headers are formatted neatly and presentably. 

## Collaborators

I collaborated with... (list names of collaborators here).


## Goals of this problem set

* Recall some exploratory data analysis skills
* Practice applying linear regression in R
* Interpret linear regression coefficients


**You are not allowed to use any generative AI models (like ChatGPT) on your problem set.**

**The Honor Principle must guide your work on this problem set.**


## Problem 0: Load packages

Load any relevant packages needed for this problem set. I added a few for you.


```{r}
library(tidyverse)
library(moderndive)
library(gglm)
```


# Part 1: Linear Models

## Problem 1: Interaction Effects and Categorical Predictors

*Based on ISLR Exercise 3.3*

We have a data set with 3 predictors, $X_1$ = GPA, $X_2$ = IQ, and $X_3$ = Level (1 for College and 0 for High School). The response variable $Y$ = Starting Salary after Graduation (in thousands of dollars). We fit the following interaction model:
$$
\hat{Y} = \beta_0 + \beta_1X_1 + \beta_2X_2 + \beta_3X_3 + \beta_4(X_1\times X_2) + \beta_5(X_1\times X_3).
$$
The estimated coefficients are as follows:
$$
\hat\beta_0 = 50, \ \hat\beta_1 = 20, \ \hat\beta_2 = 0.07, \ \hat\beta_3 = 35, \ \hat\beta_4 = 0.01, \ \hat\beta_5 = -10
$$
a.  Which of the following is correct, and why?

    1. For a fixed value of IQ and GPA, high school graduates earn more, on average, than college graduates.
    2. For a fixed value of IQ and GPA, college graduates earn more, on average, than high school graduates.
    3. For a fixed value of IQ and GPA, high school graduates earn more, on average, than college graduates provided that the GPA is high enough.
    4. For a fixed value of IQ and GPA, college graduates earn more, on average, than high school graduates provided that the GPA is high enough.

b.  Predict the salary of a college graduate with IQ of 110 and a GPA of 4.0.

c.  True or False: Since the coefficient for the GPA/IQ interaction term is very small, there is very little evidence of an interaction effect. Justify your answer.



## Problem 2: Coefficient Interpretation with Interactions

Carefully interpret each coefficient from Problem 1, in context.




## Problem 3: Nonlinear and Interaction Effects

The purpose of this problem is to engage in the model selection process by visually exploring the data you have at hand. Run the following code to load into R a dataset named `data_hw5_q5`. This dataset contains three predictor variables (`x1`, `x2`, and `x3`) and a response variable (`y`). The response variable, `y`, was truly generated by some function of these three predictors, with noise added. That function may include nonlinear and/or interaction effects!

```{r}
dat <- read_csv("simdata.csv")
dat$x3 <- factor(dat$x3)
```

a.  First, describe the data: How many observations are there? Which predictors are quantitative and which are categorical?

b.  Create the following three exploratory plots: (1) a plot displaying the distribution of the response variable, (2) a plot simultaneously displaying `x1`, `x3`, and `y`, and (3) a plot simultaneously displaying `x2`, `x3`, and `y`.

c.  Look to the second plot you created: Is there evidence of a relationship between `x1` and `y`? Does that relationship look nonlinear, and if so, what kind of nonlinearity? Is there evidence that the relationship between `x1` and `y` varies by `x3`?

d.  Look to the third plot you created: Is there evidence of a relationship between `x2` and `y`? Does that relationship look nonlinear, and if so, what kind of nonlinearity? Is there evidence that the relationship between `x2` and `y` varies by `x3`?

e.  Fit an appropriate regression model to the data. Justify your choice of model by drawing *connections to your answers to (c) and (d)* as well as the *principle of parsimony*.



## Problem 4: Multicollinearity

*Based on ISLR Exercise 3.14*

a.  Run the following code, which randomly generates values for predictors for $X_1$ and $X_2$, and then generates values for a response $Y$ based on a linear model. What are the theoretical regression coefficients for this model, as implied by the last line of code?

```{r}
set.seed(1000)
n <- 100
x1 <- runif(n)
x2 <- 0.5*x1 + rnorm(n, mean = 0, sd = 0.1)
y  <- 2 + 2*x1 + 0.3*x2 + rnorm(n, mean = 0, sd = 1 )
```

b.  What is the correlation between `x1` and `x2` (Hint: use the `cor()` function)? Create a scatterplot showing this relationship.

c.  Using the simulated data, fit a least squares regression line for $Y$ as a function of $X_1$ and $X_2$. What are the estimates for $\beta_0, \beta_1$, and $\beta_2$ (the intercept, coefficient for $X_1$, and coefficient for $X_2$, respectively)? How do these compare to the theoretical regression coefficients?

d.  Based on the regression summary, can you reject the null hypothesis $H_0: \beta_1 = 0$? What about the null hypothesis $H_0: \beta_2 = 0$? Explain.

e.  Now fit a least squares regression for $Y$ as a function of just $X_1$. Based on this model, can you reject the null hypothesis $H_0: \beta_1 = 0$?

f.  Similarly, fit a least squares regression for $Y$ as a function of just $X_2$. Based on this model, can you reject the null hypothesis $H_0: \beta_2 = 0$?

g.  Do the results in parts (d) - (f) contradict each other? Explain.



# Part 2: Accuracy Assessment and Cross Validation


## Problem 5: Concepts 

*Based on ISLR Exercise 3.4*

I collect a set of data (n = 100 observations)containing a single predictor and a quantitative response. I then fit a linear regression model to the data, as well as a separate cubic regression, i.e. $Y = \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 X^3 + \epsilon$.

a.  Suppose the true relationship between $X$ and $Y$ is linear, i.e. $Y = \beta_0 + \beta_1 X + \epsilon$. Consider the training residual sum of squares RSS for the linear regression, and also the training RSS for the cubic regression. Would we expect one to be lower than the other, would we expect them to be the same, or is there not enough information to tell? Justify your answer.

b.  Answer (a) using test rather than training RSS.

c.  Suppose that the true relationship between $X$ and $Y$ is not linear, but we don't know how far it is from linear. Consider the training RSS for the linear regression, and also the training RSS for the cubic regression. Would we expect one to be lower than the other, would we expect them to be the same, or is there not enough information to tell? Justify your answer.

d.  Answer (c) using test rather than training RSS.


## Problem 6: Hold-Out Sets

In this problem, you will develop a model to predict gas mileage based on the `Auto` data set from the `ISLR` package, which can be loaded with the following code:

```{r}
library(ISLR)
data(Auto)
```

a. Produce some numerical and graphical summaries of the `Auto` data in order to investigate the relationship between `mpg` and the other features. Discuss your findings and your process. Which variables seem most useful in predicting `mpg`? 
b. Based on your analysis in (a), fit **two** linear regression models you think will be highly predictive of `mpg`. Make sure to check diagnostic plots and adjust the models if necessary. Try to make two models that are decently different from each other.

c. Compute the training RMSEs for the models fit in (b). Which model has better predictive performance on the training data?

d. Randomly divide the data into a training and a test set using a $75 - 25$ ratio. Be sure to set a seed for reproducibility (hint: you can set a seed with the `set.seed()` function).

e. Compute the test RMSEs for the models fit in (b). Which model has better predictive performance on the testing data? 


## Problem 7: Cross Validation

In this problem, you will do further comparison of the models you fit in Problem 6. 

a. In Problem 6, you compared two models by making a 75-25 training / testing split. Discuss **at least two** advantages of using k-fold cross validation to compare these models over the 75-25 split. 

b. Describe the process of creating splits for 8-fold cross validation on this dataset. Be clear, precise, and sufficiently detailed such that someone could produce these splits with only your instructions.

c. Describe the process of fitting and predicting with the two linear models from Problem 6, on the cross validation splits. Again, be clear, precise, and sufficiently detailed such that someone could understand how to fit these models with only your instructions.

d. Finally, describe the model validation process after fitting and predicting on the cross validation splits. How do you compute a metric to decide which model is best?

e. Write your own code (not using the `tidymodels` or `rsample`) to do parts (b - d). There are many ways to do this, and not one "correct" answer. Note: there is *some* relevant code in the slides, but some of this will have to be experimentation on your part. We recommend office hours to discuss possible approaches.

f. Now, use the `tidymodels` (including `rsample`) to do parts (b - d). 

g. Report the CV RMSE for both models in parts (e) and (f) (4 values total). Did you get the exact same values for both approaches? Discuss why this does / doesn't make sense. 

h. Based on your findings, which model is best at predicting MPG? Is it **much** better than the other linear model? 

i. Finally, suppose you were given a new observation to predict MPG on based on your best model. Discuss how much you would expect your prediction to vary from the true MPG, on average. 








